Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-07 paper/ms.tex
update ms
Co-authored-by: Lynne Hillenbrand <lah@astro.caltech.edu>

Some comments on nomenclature are needed.  What we are calling
``complex periodic variables'' \citep{2023MNRAS.518.2921K} have also
been called ``complex rotators''
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
``transient flux dips'', ``persistent flux dips'', and ``scallop
shells'' \citep{2017AJ....153..152S}.  The CPVs should not be conflated
with ``dippers'', which are classical T Tauri stars with infrared
excesses, and which show large-amplitude variability linked to
obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  The phenomenology
and stellar properties of CPVs and dippers are quite different (though
see Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  The
defining phenomenological features of the CPVs are that their
light curves are {\it complex}, relative to quasi-sinusoidal
starspots, and the complex features are {\it periodic}, over
timescales of order one hundred cycles.  While rotation likely
does play a central role in explaining their physical behavior, the
acronym for ``complex rotator'' is already used in the astrophysical
literature for cosmic rays.  Given these
considerations, we refer to the stars as complex periodic variables
(CPVs); our preferred explanation for their behavior is that transient
clumps of gas or dust orbit at their corotation radii.

@@

For CPVs, binarity seems to mostly provide a mix of
nuisances and curiosities.  The
nuisances include
astrophysical false positives with two beating rapidly
rotating stars, as well as uncertainty about which
star produces the CPV signal in binary systems.  Our two candidate K dwarf CPVs 
suffer from this latter concern (see Section~\ref{subsec:resultsbinarity}).

Curiosities include the four binary systems that are now known to each host
two separate CQVs.
CQVs are sufficiently rare that such systems may have 
physical implications.
Recent work has shown that the orbits of binaries closer than
$\lesssim$700 AU tend to be aligned with their planetary systems
\citep[e.g.][]{2022AJ....163..207C}.  If we assume that observing CPV
variability requires high line-of-sight inclinations,
and that the inclinations in binaries are correlated, then we would
expect the detection of one CPV in a binary system to raise the
probability that the other star is a CPV.  The limitations of the
current catalog prevent further exploration of this issue, but it
might be interesting for future study.

A separate possible relation between binarity and CPVs is that the presence
of an intermediate-separation binary could cause early disk dispersal,
freeing the star to contract.  The CPV phenomenon seems to require
rapid rotation; thus if binary systems tend to produce more rapidly
rotating stars, we would expect a sample of CPVs to have a larger
binary fraction than the field.  However, we see minimal evidence for
such an effect (Section~\ref{subsec:resultsbinarity}).  The
multiplicity rate of M dwarfs near the Sun is 26.8$\pm$1.4\%
\citep{2019AJ....157..216W}.  Based on the same study, the peak of the
separation distribution decreases from $\approx$49\,AU for
$0.30$-$0.60$\,$M_\odot$ stars, to $\approx$11\,AU for
$0.15$-$0.30$\,$M_\odot$ M dwarfs.  The multiplicity fraction in our
CPV sample ($\approx$32\% from RUWE$_{\rm DR3}$) seems approximately
consistent with this field fraction.

\subsection{Transience \& special phases of CPV dips}

While CPV periods appear to remain fixed over thousands of cycles, the light curve shapes evolve over
typical timescales of 10 to 1{,}000 cycles (e.g.
Figures~\ref{fig:evoln} and~\ref{fig:lpriver0}).  Although we refer
to them as ``periodic'', the CPVs are
therefore actually quasiperiodic, with quality factors of $Q$$\approx$100.
While this observation is consistent with the analyses by
\citet{2022AJ....163..144G} and \citet{2023ApJ...945..114P}, the fact
that all CPV dips seem to be transient over timescales of more than
1{,}000 cycles (Figure~\ref{fig:evoln}) marks a qualitative departure
from the ``persistent'' {\it vs.} ``transient'' flux dip distinction
previously described by \citet{2017AJ....153..152S}.

One peculiarity of CPV evolution however is that the dips do not
explore all phase angles with equal weight.  LP 12-502, and other CPVs,
have preferred phases lasting for at least two years.  For LP 12-502,
all of the dips happen over phases corresponding to only two thirds of
the period (Figures~\ref{fig:lp} and~\ref{fig:lpriver0}).  The
remaining third seems to be ``out of limits'' for dips over the
timespan of observations.  This could be evidence that the stellar
magnetic field is not azimuthally symmetric.  Alternatively, the
source of the material (e.g. a planetesimal swarm) might be
distributed over an arc rather than occurring randomly around the
entire orbit.

@@

Examining Figure~\ref{fig:cqvs}, it is clear that CPV dips can be
asymmetric but it is not obvious whether there is a preference for
sharper ingresses or sharper egresses.  In some cases (e.g.
TIC~425933644), the flux variations do not resemble isolated dips,
making the meaning of ``ingress'' and ``egress'' unclear.  In other
cases, such as Sector~36 of TIC~89463560, there is a sharp drop with
an exponential return to the baseline flux, resembling the signatures
of exocomet \citep[e.g.][]{2018MNRAS.474.1453R,2019A&A...625L..13Z},
or the outflowing exospheres of certain transiting planets
\citep[e.g.][]{2019ApJ...873...89M,2022ApJ...926..226M}.

@@

In the context of disk-driven planet migration, the stopping location
for the innermost planet is set by the location of the protoplanetary
disk's truncation radius \citep[e.g.][and references therein]{2018haex.bookE.142I}.  The 
truncation radius is traditionally calculated by equating the magnetic
pressure from the stellar magnetosphere with the ram pressure of the
inflowing gas.  As it happens, the truncation radius is close to the
corotation radius for low accretion rates
\citep[e.g.][]{2015SSRv..191..339R,2022MNRAS.510.5246L}.  These considerations
invite us to imagine one or more planets migrating inward due to gas
drag, and arriving at $\approx$5-10 stellar
radii before the disk is depleted.

>>>>

Some comments on nomenclature are needed.  What we are calling
``complex periodic variables'' \citep{2023MNRAS.518.2921K} have also
been called ``complex rotators''
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
``transient flux dips'', ``persistent flux dips'', and ``scallop
shells'' \citep{2017AJ....153..152S}.  The CPVs should not be
conflated with ``dippers'', which are classical T Tauri stars with
infrared excesses, and which show large-amplitude variability linked
to obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  The phenomenology
and stellar properties of CPVs and dippers are quite different (though
see Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  The
defining phenomenological features of the CPVs are that their light
curves are {\it complex}, relative to quasi-sinusoidal starspots, and
the complex features are {\it periodic}, meaning they typically repeat
for at least tens of days.  While rotation likely does play a central
role in explaining their physical behavior, the acronym for ``complex
rotator'' is already used in the astrophysical literature for cosmic
rays.  Given these considerations, we refer to the stars as complex
periodic variables (CPVs); our preferred explanation for their
behavior is that transient clumps of gas or dust orbit at their
corotation radii.

@@

For CPVs, binarity seems to provide either nuisances or curiosities.  The
nuisances include
astrophysical false positives with two beating rapidly
rotating stars, as well as uncertainty about which
star produces the CPV signal in binary systems.  Our two candidate K dwarf CPVs 
suffer from this latter concern (see Section~\ref{subsec:resultsbinarity}).

Curiosities include the four binary systems that are now known to each host
two separate CQVs.
CQVs are sufficiently rare that such systems may have 
physical import.
Recent work has shown that the orbits of binaries closer than
$\lesssim$700 AU tend to be aligned with their planetary systems
\citep[e.g.][]{2022AJ....163..207C}.  If we assume that observing CPV
variability requires high line-of-sight inclinations,
and that the inclinations in binaries are correlated, then we would
expect the detection of one CPV in a binary system to raise the
probability that the other star is a CPV.  The limitations of the
current catalog prevent further exploration of this issue, but it
might be interesting for future study.

A separate possible relation between binarity and CPVs is that the presence
of a binary companion could cause early disk dispersal,
freeing the star to contract.  The CPV phenomenon seems to require
rapid rotation; thus if binary systems tend to produce more rapidly
rotating stars, we would expect a sample of CPVs to have a larger
binary fraction than the field.  However, we see minimal evidence for
such an effect (Section~\ref{subsec:resultsbinarity}).  The
multiplicity rate of M dwarfs near the Sun is 26.8$\pm$1.4\%
\citep{2019AJ....157..216W}.  
%Based on the same study, the peak of the
%separation distribution decreases from $\approx$49\,AU for
%$0.30$-$0.60$\,$M_\odot$ stars, to $\approx$11\,AU for
%$0.15$-$0.30$\,$M_\odot$ M dwarfs.
The $\approx$32\% multiplicity fraction in our
CPV sample (from RUWE$_{\rm DR3}$; see Section~\ref{subsec:resultsbinarity})
seems approximately
consistent with this field fraction.

\subsection{Transience of CPV dips}

While CPV periods appear to remain fixed over thousands of cycles, the light curve shapes evolve over
typical timescales of 10 to 1{,}000 cycles (e.g.\ Figures~\ref{fig:evoln} and~\ref{fig:lpriver0}).  Although we refer
to them as ``periodic'', the CPVs are
therefore actually quasiperiodic, with coherence timescales of $\approx$100 cycles.
This marks a qualitative departure
from the ``persistent'' {\it vs.} ``transient'' flux dip distinction
previously described by \citet{2017AJ....153..152S}, which was based on
$\lesssim$100\,cycle K2 baselines.
The observation that CPVs have a population-averaged on-off duty cycle of $\approx$75\%
(Figure~\ref{fig:evoln}) is also new.
Appendix~\ref{app:tic3006} for instance shows $\approx$1{,}000 cycles of a source, TIC~300651846, with between zero and five sharp local minima per cycle.  During the ``zero'' epochs (cycles $\approx$503-542), the source would likely be labeled an ordinary rotating star.

\subsection{Special phases of CPV dips}

An independent peculiarity of CPV evolution is that the dips do not
explore all phase angles with equal weight.  LP 12-502, and other CPVs,
exhibit preferred phases lasting for at least two years.  For LP 12-502,
all of the dips happen over phases corresponding to only two thirds of
the period (Figures~\ref{fig:lp} and~\ref{fig:lpriver0}).  The
remaining third seems to be ``out of limits'' for dips over the
timespan of observations.  This could be evidence that the stellar
magnetic field is not azimuthally symmetric.  Alternatively, the
source of the material (e.g. a planetesimal swarm) might be
distributed over an arc rather than occurring randomly around the
entire orbit.

@@

Examining Figure~\ref{fig:cqvs}, it is clear that CPV dips can be
asymmetric but it is not obvious whether there is a preference for
sharper ingresses or sharper egresses.  In some cases (e.g.
TIC~425933644), the flux variations do not resemble isolated dips,
making the meaning of ``ingress'' and ``egress'' unclear.  In other
cases, such as Sector~36 of TIC~89463560, there is a sharp drop with
an exponential return to the baseline flux, resembling the signatures
of exocomet \citep[e.g.][]{2018MNRAS.474.1453R,2019A&A...625L..13Z},
and the outflowing exospheres of some transiting planets
\citep[e.g.][]{2019ApJ...873...89M,2022ApJ...926..226M}.

@@

In the context of disk-driven planet migration, the stopping location
for the innermost planet is set by the protoplanetary
disk's truncation radius \citep[e.g.][and references therein]{2018haex.bookE.142I}.  The 
truncation radius is often calculated by equating the magnetic
pressure from the stellar magnetosphere with the ram pressure of the
inflowing gas.  As it happens, the truncation radius is close to the
corotation radius for low accretion rates
\citep[e.g.][]{2015SSRv..191..339R,2022MNRAS.510.5246L}.  These considerations
invite us to imagine one or more planets migrating inward due to gas
drag, and arriving at $\approx$5-10 stellar
radii before the disk is depleted.
